---
title: "Continuous- and discrete-time deterministic and probabilistic models"
format: html
editor: visual
---

## Packages

```{r}
library(deSolve)
library(purrr)
library(adaptivetau)
```

## Functions

```{r}
plot2 <- function(...) plot(..., lwd = 2, col = 4)
```

```{r}
lines2 <- function(...) lines(..., lwd = 2)
```

```{r}
replicate2 <- function(...) replicate(..., simplify = FALSE)
```


## Death process

### Continuous-time deterministic

$$
\frac{dN}{dt} = -r_cN
$$

```{r}
continuous_deterministic <- function(N0, rc, times) {
  tibble::as_tibble(
    as.data.frame(
      ode(c(N = N0),
          times,
          function(time, variables, parameters) {
            with(as.list(c(variables, parameters)), {
              dN <- rc * N
              list(dN)
            })
          },
          c(rc = rc))
    )
  )
}
```


```{r}
continuous_deterministic(1000, -.2, seq(0, 10, .1)) |> 
  with(plot2(time, N, type = "l"))
```


### Discrete-time deterministic

$$
N_{t+1} = r_d N_t
$$

with

$$
r_d = e^{r_c \Delta t}
$$

$\Delta t$ being the duration of the time step.

```{r}
discrete_deterministic <- function(N0, rc, times) {
  rd <- exp(rc * mean(diff(times)))
  N <- N0
  for (i in times[-1]) {
    N <- c(N, tail(N, 1) * rd)
  }
  tibble::tibble(time = times, N = N)
}
```

```{r}
compare_continuous_discrete_deterministic <- function(
    step, N0 = 1000, rc = -.2, tmin = 0, tmax = 10) {
  
  continuous_deterministic(N0, rc, seq(tmin, tmax, step)) |> 
    with(plot2(time, N, type = "l"))
  
  discrete_deterministic(N0, rc, seq(tmin, tmax, step)) |> 
    with(points(time, N, lwd = 2, col = 2))
}
```

```{r}
compare_continuous_discrete_deterministic(.1)
compare_continuous_discrete_deterministic(1)
```

### Discrete-time probabilistic

$$
p = 1 - e^{r_c\Delta t}
$$

```{r}
discrete_probabilistic <- function(N0, rc, times) {
  Delta_t <- mean(diff(times))
  p <- 1 - exp(rc * Delta_t)
  N <- N0
  for (i in times[-1]) {
    N <- c(N, tail(N, 1) * (1 - p))
  }
  tibble::tibble(time = times, N = N)
}
```

```{r}
compare_discrete_deterministic_probabilistic <- function(
    step, N0 = 1000, rc = -.2, tmin = 0, tmax = 10) {
  
  det <- discrete_deterministic(N0, rc, seq(tmin, tmax, step))
  prb <- discrete_probabilistic(N0, rc, seq(tmin, tmax, step))
  plot(det$N, prb$N, col = 4, lwd = 2)
  abline(0, 1)
}
```

```{r}
compare_discrete_deterministic_probabilistic(.1)
compare_discrete_deterministic_probabilistic(1)
```

### Discrete-time stochastic

```{r}
discrete_stochastic <- function(N0, rc, times) {
  Delta_t <- mean(diff(times))
  p <- 1 - exp(rc * Delta_t)
  N <- N0
  for (i in times[-1]) {
    tailN <- tail(N, 1)
    N <- c(N, tailN - rbinom(1, tailN, p))
  }
  tibble::tibble(time = times, N = N)
}
```

```{r}
out_d <- replicate2(1000, discrete_stochastic(1000, -.2, seq(0, 10, 1)))
```

```{r}
plot(NA, xlab = "time", ylab = "N", xlim = c(0, 10), ylim = c(0, 1000))
walk(out_d, ~ with(.x, points(time, N, type = "o", col = adjustcolor(4, .01))))
```

### Continuous-time stochastic

```{r}
continuous_stochastic <- function(N0, rc, times) {
  N <- N0
  tf <- max(times)
  times <- min(times)
  while (tail(times, 1) < tf & N > -1) {
    times <- c(times, tail(times, 1) + rexp(1, N * abs(rc)))
    N <- N - 1
  }
  times <- head(times, -1)
  tibble::tibble(time = times, N = N0 - seq_along(times))
}
```

```{r}
out_c <- continuous_stochastic(1000, -.2, 0:10)
```

```{r}
with(out_c, plot(time, N, type = "s"))
```

```{r}
out_c <- replicate2(1000, continuous_stochastic(1000, .2, 0:10))
```

```{r}
plot(NA, xlab = "time", ylab = "N", xlim = c(0, 10), ylim = c(0, 1000))
walk(out_c, ~ with(.x, points(time, N, type = "s")))
walk(out_d, ~ with(.x, points(time, N, col = 4)))
```

## Birth process

### Continuous-time deterministic

```{r}
continuous_deterministic(1000, .2, seq(0, 10, .1)) |> 
  with(plot2(time, N, type = "l"))
```
### Discrete-time deterministic

```{r}
compare_continuous_discrete_deterministic(.1, rc = .2)
```

```{r}
compare_continuous_discrete_deterministic(1, rc = .2)
```

### Discrete-time probabilistic



```{r}
compare_discrete_deterministic_probabilistic(.1, rc = .2)
```

```{r}
compare_discrete_deterministic_probabilistic(1, rc = .2)
```

### Discrete-time stochastic

```{r}
out_d <- replicate2(1000, discrete_stochastic(1000, -.2, seq(0, 10, 1)))
plot(NA, xlab = "time", ylab = "N", xlim = c(0, 10), ylim = c(0, 1000))
walk(out_d, ~ with(.x, points(time, N, type = "o", col = adjustcolor(4, .01))))
```


